---
title: "Lab Session: Trees, Boosting, Bagging, and Random Forests"
date: "Class 10"
output:
  html_document:
    toc: FALSE
    toc_float: true
    theme: united
---

```{=html}
<style type="text/css">
    .main-container {
      max-width: 1600px !important; /* Adjust this value as needed for the main content */
    }

    /* Optional: Adjust the width of the floating TOC */
    .tocify {
      width: 200px; /* Adjust this value to control TOC width */
    }

    /* Optional: Ensure content doesn't overlap with TOC on scroll */
    @media (min-width: 768px) { /* Apply for larger screens where TOC floats */
      body.toc-left .main-container {
        margin-left: 210px; /* Adjust this to be slightly more than TOC width */
      }
    }
    body {
  font-size: 12pt; /* Sets the base font size for the entire document */
}
h1,h2,h3,h4,h5,h6{
  font-size: 24pt;
}
code {
  font-size: 12pt; /* Sets font size for code blocks */
}
output {
  font-size: 24pt; /* Sets font size for code blocks */
}

</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center", fig.width = 12, fig.height = 10)
```

## Lab: Tree based methods using Heart Data {.tabset}

In this lab, we will explore tree-based methods for classification. We'll use the `Heart` dataset from the `ISLR2` package, which contains a binary outcome `AHD` for 303 patients who presented with chest pain. An outcome value of `Yes` indicates the presence of heart disease based on an angiographic test, while `No` means no heart disease. There are $p=13$ predictors including `Age`, `Sex`, `Chol` (a cholesterol measurement), and other heart and lung function measurements. 

```{r load-packages}
rm(list = ls())
pacman::p_unload(pacman::p_loaded())
pacman::p_load(tidyverse, ISLR2, tree, patchwork, ggsci, caret, randomForest, gbm)

### section 1: basic EDA
heart <- read_csv("https://www.statlearning.com/s/Heart.csv") |> 
  select(-1) |> 
  drop_na() |> 
  mutate(across(where(is.character), as.factor))
```

---

### Data, EDA, & Preprocessing

First, we load the `Heart` dataset. 

```{r}
str(heart)
summary(heart)
```

#### Train/Test Split

We need to evaluate model performance on unseen data. We'll split the data into a training set (4/5) and a test set (1/5)

```{r, echo = FALSE}
### section 1: split into test and train ###
set.seed(42)
train_index <- sample(1:nrow(heart), replace = FALSE)[1:round(0.8*nrow(heart))]
test_index <- (1:nrow(heart))[-train_index]
cat("Training set size:", length(train_index), "\n")
cat("Test set size:", length(test_index), "\n")
```




### Basic tree building and pruning


```{r}
model <- tree(AHD ~ ., data = heart[train_index,])
plot(model)
text(model, pretty = 0)

set.seed(42)
cv_results <- cv.tree(model, FUN = prune.misclass)
best <- cv_results$size[which.min(cv_results$dev)]

pruned_model <- prune.misclass(model, best = best)
plot(pruned_model)
text(pruned_model, pretty = 0)
```

```{r}
caret::confusionMatrix(data = predict(model, newdata = heart[test_index,], type = "class"), 
                       reference = heart[test_index,]$AHD)
```

```{r}
caret::confusionMatrix(data = predict(pruned_model, newdata = heart[test_index,], type = "class"), 
                       reference = heart[test_index,]$AHD)
```

### Bagging


```{r}
set.seed(42)
model_bagged <- randomForest::randomForest(AHD ~., data = heart[train_index,], mtry = ncol(heart) - 1)
caret::confusionMatrix(data = predict(model_bagged, newdata = heart[test_index,], type = "class"), 
                       reference = heart[test_index,]$AHD)
importance(model_bagged, type = 1)
```

### Boosting 

```{r}
### section 4: boosting
set.seed(42)
model_boosted <- gbm(AHD ~., data = heart[train_index,] |> mutate(AHD = case_when(AHD == "Yes" ~ 1, .default = 0)), 
                     distribution = "bernoulli", interaction.depth = 4, n.trees = 1000, shrinkage = 0.2,
                     verbose = F)
caret::confusionMatrix(data = as.factor(ifelse(predict(model_boosted, newdata = heart[test_index,], type = "response") <= 0.5, 
                                     "No", "Yes")), 
                       reference = heart[test_index,]$AHD)
summary(model_boosted, plotit = FALSE)
```

### Random forest

```{r}
### section 5 random forest
model_rf <- randomForest::randomForest(AHD ~., data = heart[train_index,], mtry = sqrt(ncol(heart) - 1), importance = TRUE)
caret::confusionMatrix(data = predict(model_rf, newdata = heart[test_index,], type = "class"), 
                       reference = heart[test_index,]$AHD)
importance(model_rf, type = 1)
```
