---
title: "Class 3/Lab1: Subset Selection and Shrinkage"
date: "Week 3 (9.11)"
output:
  html_document:
    toc: FALSE
    toc_float: true
    theme: cosmo
---

```{=html}
<style type="text/css">
    .main-container {
      max-width: 1600px !important; /* Adjust this value as needed for the main content */
    }

    /* Optional: Adjust the width of the floating TOC */
    .tocify {
      width: 200px; /* Adjust this value to control TOC width */
    }

    /* Optional: Ensure content doesn't overlap with TOC on scroll */
    @media (min-width: 768px) { /* Apply for larger screens where TOC floats */
      body.toc-left .main-container {
        margin-left: 210px; /* Adjust this to be slightly more than TOC width */
      }
    }

</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

## Introduction {.tabset}

In this lab, we will:

1.  Fit a simple linear regression baseline and interpret its coefficients and prediction intervals.
2.  Compare best subset, forward stepwise, and backward stepwise selection on the `insurance` dataset.
3.  Fit ridge and lasso regression, and visualize how their coefficient paths differ as the penalty grows.

```{r load-packages}
# Clear environment and unload packages
rm(list = ls())
pacman::p_unload(pacman::p_loaded())

# Load necessary packages
pacman::p_load(tidyverse, leaps, glmnet, patchwork, here, latex2exp)
```

---

### Simple Linear Regression Baseline

Before comparing selection methods, it's worth fitting the simplest possible model: one predictor, ordinary least squares. We'll use the `neonatal_brain_volumes` dataset to relate gestational age (`GA`) to brain volume.

```{r slr-data}
brain_data <- read_csv(file.path(here(), "lectures", "03-regression-1", "data", "neonatal_brain_volumes.csv"))
head(brain_data)
```

```{r slr-fit}
slr_fit <- lm(`brain volume` ~ GA, data = brain_data)
summary(slr_fit)
```

```{r slr-plot, fig.width=6, fig.height=5}
ggplot(brain_data, aes(x = GA, y = `brain volume`)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_smooth(method = "lm", color = "#dc3545", fill = "#dc3545", alpha = 0.15) +
  labs(title = "Brain Volume vs. Gestational Age", x = "Gestational Age (weeks)", y = "Brain Volume") +
  theme_minimal(base_size = 14)
```

```{r slr-intervals}
confint(slr_fit, parm = "GA")

new_ga <- data.frame(GA = mean(brain_data$GA))
predict(slr_fit, newdata = new_ga, interval = "confidence")
predict(slr_fit, newdata = new_ga, interval = "prediction")
```

**Interpretation:** The confidence interval describes uncertainty about the *mean* brain volume at a given gestational age; the (wider) prediction interval describes uncertainty about a single *new* observation. Every method below is, at its core, still fitting a linear model like this one — just with a strategy for choosing which predictors belong in it.

---

### Subset Selection

With more than a couple of predictors, we need a systematic way to choose which ones to include. We'll compare three approaches on the `insurance` dataset (predicting `charges` from the other columns).

```{r subset-data}
insurance <- read_csv(file.path(here(), "lectures", "03-regression-1", "data", "insurance.csv"))
head(insurance)
```

```{r subset-fit}
subset_formula <- charges ~ .

fit_best     <- regsubsets(subset_formula, data = insurance, nvmax = ncol(insurance) - 1)
fit_forward  <- regsubsets(subset_formula, data = insurance, nvmax = ncol(insurance) - 1, method = "forward")
fit_backward <- regsubsets(subset_formula, data = insurance, nvmax = ncol(insurance) - 1, method = "backward")
```

#### Which variables enter at each model size?

```{r subset-outmat}
summary(fit_best)$outmat
```

Each row is a model size; each `*` marks a predictor included at that size. Compare this to `summary(fit_forward)$outmat` and `summary(fit_backward)$outmat` — for a small predictor set like this one, the three methods often agree, but that's not guaranteed in general (forward and backward selection are greedy and can miss the best subset that exhaustive search would find).

#### Choosing a model size

`regsubsets` reports RSS, adjusted $R^2$, and Mallows' $C_p$ for every model size; we compute BIC ourselves from the selected variables at each step.

```{r subset-criteria}
calculate_bic <- function(model_obj, data) {
  s <- summary(model_obj)
  sapply(seq_len(nrow(s$outmat)), function(i) {
    vars <- names(which(s$outmat[i, ] == "*"))
    vars <- vars[vars != "(Intercept)"]
    fit <- lm(reformulate(vars, response = "charges"), data = data)
    BIC(fit)
  })
}

criteria_df <- data.frame(
  num_vars = seq_along(summary(fit_best)$rss),
  rss      = summary(fit_best)$rss,
  adjr2    = summary(fit_best)$adjr2,
  cp       = summary(fit_best)$cp,
  bic      = calculate_bic(fit_best, insurance)
)

criteria_df
```

```{r subset-criteria-plot, fig.width=9, fig.height=6}
p_adjr2 <- ggplot(criteria_df, aes(num_vars, adjr2)) +
  geom_line(color = "steelblue") + geom_point(color = "steelblue") +
  geom_point(data = criteria_df[which.max(criteria_df$adjr2), ], color = "red", size = 3) +
  labs(x = "Number of Variables", y = "Adjusted R²") + theme_bw()

p_cp <- ggplot(criteria_df, aes(num_vars, cp)) +
  geom_line(color = "steelblue") + geom_point(color = "steelblue") +
  geom_point(data = criteria_df[which.min(criteria_df$cp), ], color = "red", size = 3) +
  labs(x = "Number of Variables", y = "Mallows' Cp") + theme_bw()

p_bic <- ggplot(criteria_df, aes(num_vars, bic)) +
  geom_line(color = "steelblue") + geom_point(color = "steelblue") +
  geom_point(data = criteria_df[which.min(criteria_df$bic), ], color = "red", size = 3) +
  labs(x = "Number of Variables", y = "BIC") + theme_bw()

p_rss <- ggplot(criteria_df, aes(num_vars, rss)) +
  geom_line(color = "steelblue") + geom_point(color = "steelblue") +
  geom_point(data = criteria_df[which.min(criteria_df$rss), ], color = "red", size = 3) +
  labs(x = "Number of Variables", y = "RSS") + theme_bw()

(p_rss + p_adjr2) / (p_cp + p_bic) +
  plot_annotation(title = "Model Selection Criteria by Subset Size", subtitle = "Red point marks the optimum for each criterion")
```

**Interpretation:** RSS and (in-sample) $R^2$ always improve as you add predictors, so they can't be used to *choose* a model size — that's exactly why $C_p$, BIC, and adjusted $R^2$ exist: each penalizes complexity differently, and BIC penalizes it most heavily. Where the red dots land at different sizes across panels is itself the lesson: the "best" model size depends on which tradeoff between fit and parsimony you care about.

---

### Ridge and Lasso Regression

Subset selection chooses a hard in/out for each predictor. Ridge and lasso instead *shrink* coefficients continuously, trading a little bias for a lot less variance. We'll fit both on the same `insurance` data, using `age` and `bmi` as predictors of `charges`.

```{r shrink-data}
x_vars <- insurance %>% select(age, bmi) %>% scale()
y_var  <- insurance$charges

ols_fit <- lm(charges ~ age + bmi, data = insurance %>% mutate(age = as.numeric(x_vars[, "age"]), bmi = as.numeric(x_vars[, "bmi"])))
coef(ols_fit)
```

```{r shrink-fit}
ridge_fit <- glmnet(x_vars, y_var, alpha = 0)
lasso_fit <- glmnet(x_vars, y_var, alpha = 1)
```

#### Coefficient paths vs. $\lambda$

As $\lambda$ grows, ridge coefficients shrink smoothly toward zero but rarely hit it exactly; lasso coefficients can be driven to exactly zero, which is what makes it useful for variable selection.

```{r shrink-path-lambda, fig.width=10, fig.height=5}
tidy_path <- function(fit, model_name) {
  as.matrix(coef(fit)) %>%
    as.data.frame() %>%
    rownames_to_column("Variable") %>%
    pivot_longer(-Variable, names_to = "step", values_to = "Coefficient") %>%
    mutate(Lambda = fit$lambda[as.numeric(str_remove(step, "s")) + 1],
           Model = model_name) %>%
    filter(Variable != "(Intercept)")
}

path_df <- bind_rows(tidy_path(ridge_fit, "Ridge"), tidy_path(lasso_fit, "Lasso"))

ggplot(path_df, aes(x = Lambda, y = Coefficient, color = Variable)) +
  geom_line(linewidth = 1) +
  scale_x_log10() +
  facet_wrap(~Model, scales = "free_x") +
  labs(title = "Coefficient Path vs. Lambda", x = TeX("$\\lambda$ (log scale)"), y = "Coefficient") +
  theme_bw(base_size = 13)
```

#### Choosing $\lambda$ by cross-validation

```{r shrink-cv, fig.width=10, fig.height=5}
set.seed(2155)
cv_ridge <- cv.glmnet(x_vars, y_var, alpha = 0)
cv_lasso <- cv.glmnet(x_vars, y_var, alpha = 1)

par(mfrow = c(1, 2))
plot(cv_ridge); title("Ridge", line = 2.5)
plot(cv_lasso); title("Lasso", line = 2.5)
par(mfrow = c(1, 1))
```

```{r shrink-coefs}
cat("Ridge coefficients at lambda.min:\n")
coef(cv_ridge, s = "lambda.min")

cat("\nLasso coefficients at lambda.min:\n")
coef(cv_lasso, s = "lambda.min")
```

**Interpretation:** Cross-validation picks the $\lambda$ that minimizes out-of-sample error, marked by the left dashed line in each plot (`lambda.min`); the right dashed line (`lambda.1se`) gives the most parsimonious model within one standard error of the minimum — a common choice when you'd rather have a simpler model at a small cost in accuracy. Note that with only two predictors, lasso may not zero anything out here; its selection behavior becomes more visible as the predictor set grows.

#### Comparing paths side by side

```{r shrink-compare, fig.width=7, fig.height=6}
ridge_coefs <- as.vector(coef(cv_ridge, s = "lambda.min"))[-1]
lasso_coefs <- as.vector(coef(cv_lasso, s = "lambda.min"))[-1]
ols_coefs   <- coef(ols_fit)[-1]

xlim <- range(c(ridge_coefs, lasso_coefs, ols_coefs, 0)) + c(-500, 500)
ylim <- xlim

plot(0, 0, type = "n", xlim = xlim, ylim = ylim,
     xlab = "Coefficient (age, scaled)", ylab = "Coefficient (bmi, scaled)",
     main = "OLS vs. Ridge vs. Lasso Estimates")
grid()
abline(h = 0, lty = 3, col = "gray50"); abline(v = 0, lty = 3, col = "gray50")

points(ols_coefs[1], ols_coefs[2], pch = 19, cex = 2, col = "black")
points(ridge_coefs[1], ridge_coefs[2], pch = 8, cex = 2, col = "red")
points(lasso_coefs[1], lasso_coefs[2], pch = 17, cex = 2, col = "darkgreen")

legend("topleft", legend = c("OLS", "Ridge (CV)", "Lasso (CV)"),
       col = c("black", "red", "darkgreen"), pch = c(19, 8, 17), pt.cex = 1.5, bty = "n")
```

**Interpretation:** Ridge and lasso estimates sit closer to the origin than the unpenalized OLS estimate — both are trading a small amount of bias for reduced variance. The shape of the constraint region each method implicitly enforces (a circle for ridge, a diamond for lasso) is what makes lasso capable of setting coefficients to exactly zero, since the diamond's corners lie on the axes.
