---
title: "Lab Session: Unsupervised Learning - Clustering"
date: "Class 8"
output:
  html_document:
    toc: FALSE
    toc_float: true
    theme: united
---

```{=html}
<style type="text/css">
    .main-container {
      max-width: 1600px !important; /* Adjust this value as needed for the main content */
    }

    /* Optional: Adjust the width of the floating TOC */
    .tocify {
      width: 200px; /* Adjust this value to control TOC width */
    }

    /* Optional: Ensure content doesn't overlap with TOC on scroll */
    @media (min-width: 768px) { /* Apply for larger screens where TOC floats */
      body.toc-left .main-container {
        margin-left: 210px; /* Adjust this to be slightly more than TOC width */
      }
    }

</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

## Lab: K-Means and Hierarchical Clustering {.tabset}

In this lab, we will explore two common unsupervised clustering methods: Hierarchical Clustering and K-Means. Our goal is to discover natural groupings within a dataset of European countries based on their employment percentages in different industries.

```{r load-packages}
rm(list = ls())
pacman::p_unload(pacman::p_loaded())
pacman::p_load(tidyverse, cluster, factoextra, here, knitr)
```

---

### Data, EDA, & Scaling

First, we'll load the `Eurojobs` dataset. This dataset contains the percentage of the workforce employed in nine different industries for 26 European countries in 1979. It contains 10 variables:

- `Country`: the name of the country (identifier)
- `Agr`: % of workforce employed in agriculture
- `Min`: % in mining
- `Man`: % in manufacturing
- `PS`: % in power supplies industries
- `Con`: % in construction
- `SI`: % in service industries
- `Fin`: % in finance
- `SPS`: % in social and personal services
- `TC`: % in transportation and communications

We will perform some basic exploratory data analysis (EDA) and then—most importantly—scale the data. Scaling is crucial for distance-based algorithms like K-Means and Hierarchical Clustering, as it ensures that variables with larger ranges (e.g., Agriculture) don't dominate the clustering process over variables with smaller ranges (e.g., Mining).

```{r load-data}
# Load the data from the URL provided in the inspiration HTML file
data_url <- "https://statsandr.com/blog/data/Eurojobs.csv"
Eurojobs <- read.csv(data_url, row.names = 1) # Use first column as row names
print(summary(Eurojobs))
```

Now, we scale the data. We'll use the `scale()` function, which standardizes each column to have a mean of 0 and a standard deviation of 1.

```{r scale-data}
# scale the data
Eurojobs_scaled <- scale(Eurojobs)
print(summary(as.data.frame(Eurojobs_scaled)))
```

---

### Hierarchical Clustering

Hierarchical (or agglomerative) clustering builds a "bottom-up" hierarchy of clusters. It starts with each observation as its own cluster and progressively merges the closest pairs of clusters until all observations are in a single cluster.

We will:
1.  Calculate a distance matrix for our scaled data.
2.  Run the hierarchical clustering algorithm (using "complete" linkage).
3.  Plot the resulting dendrogram to visualize the cluster hierarchy.

```{r hc-dist}
dist_matrix <- dist(Eurojobs_scaled, method = "euclidean")
```

```{r hc-cluster}
# other methods include "average", "single", and "centroid"
hc_complete <- hclust(dist_matrix, method = "complete")

# plot the dendrogram - basic
# plot(hc_complete, main = "Complete Linkage Dendrogram", xlab = "", sub = "")

# fviz_dend from factoextra gives a nicer, clearer plot
fviz_dend(hc_complete, 
          main = "Dendrogram (Complete Linkage)",
          cex = 0.7, # Label size
          k = 4, # Cut the tree into 4 clusters
          rect = TRUE, # Add rectangles around the clusters
          rect_border = "jco",
          rect_fill = TRUE,
          palette = "jco",
          ggtheme = theme_minimal())
```

Looking at the dendrogram, we can decide where to "cut" the tree to form clusters. For example, a cut at a height of ~5.0 would give us 4 distinct clusters. We can get the cluster assignments for each country using the `cutree()` function.

```{r hc-cut}
# cut the tree to get 4 clusters
clusters_hc <- cutree(hc_complete, k = 4)

# Show which countries are in which cluster
Eurojobs_clustered_hc <- Eurojobs %>%
  mutate(Cluster = as.factor(clusters_hc))

cat("\n--- Countries by Cluster (Hierarchical) ---\n")
# Print countries for each cluster
for(i in 1:4) {
  cat(paste("\nCluster", i, ":\n"))
  print(rownames(Eurojobs_clustered_hc[Eurojobs_clustered_hc$Cluster == i, ]))
}
```

---

### K-Means & Gap Statistic

K-Means clustering is a "top-down" approach where we must specify the number of clusters, `k`, in advance. The algorithm then partitions the data into `k` clusters, minimizing the within-cluster sum of squares.

The main challenge is choosing the right `k`. We will use three methods to find the optimal `k`, with a special focus on the **Gap Statistic** as requested.


1.  **Silhouette Method:** Measures how similar an observation is to its own cluster compared to other clusters. We look for the `k` with the highest average silhouette score.
2.  **Gap Statistic:** Compares the WSS of our data to the WSS of randomly generated "null" reference datasets. We pick the `k` that maximizes this "gap."

```{r find-k, fig.width=12, fig.height=4}
# seed for reproducible results
set.seed(42)
# silhouette Method
p2 <- fviz_nbclust(Eurojobs_scaled, kmeans, method = "silhouette") +
  labs(subtitle = "Silhouette Method")

# gap Statistic Method
# We use nboot = 50 to speed it up for the lab. 
# For a real analysis, nboot = 500 or more is better.
p3 <- fviz_nbclust(Eurojobs_scaled, kmeans, 
                   method = "gap_stat", 
                   nboot = 50) +
  labs(subtitle = "Gap Statistic")

# Display plots side-by-side
gridExtra::grid.arrange(p2, p3, nrow = 1)
```

**Interpretation:**

* **Silhouette:** Suggests `k=` `r which.max(p2$data$y)` as the optimal number of clusters.
* **Gap Statistic:** Clearly shows the gap statistic peaking at `k=` `r which.max(p3$data$gap)`.

Now, let's run our final K-Means algorithm with `k=` `r which.max(p3$data$gap)`. 

```{r kmeans-final, fig.width=8, fig.height=6}
# Set seed for reproducibility
set.seed(42)
# Perform k-means clustering with k = 3
km_final <- kmeans(Eurojobs_scaled, centers = which.max(p3$data$gap))
```

We can inspect the cluster centers (the average *scaled* values for each industry) and see which countries were assigned to which cluster.

```{r kmeans-results}
# add K-Means cluster assignments to the original data
Eurojobs_clustered_km <- Eurojobs %>%
  mutate(Cluster = as.factor(km_final$cluster))

cat("\n--- Countries by Cluster (K-Means) ---\n")
# Print countries for each cluster
for(i in 1:which.max(p3$data$gap)) {
  cat(paste("\nCluster", i, ":\n"))
  print(rownames(Eurojobs_clustered_km[Eurojobs_clustered_km$Cluster == i, ]))
}
```
