---
title: "Lab Session: Model averaging"
date: "Class 11"
output:
  html_document:
    toc: FALSE
    toc_float: true
    theme: united
---

```{=html}
<style type="text/css">
    .main-container {
      max-width: 1600px !important; /* Adjust this value as needed for the main content */
    }

    /* Optional: Adjust the width of the floating TOC */
    .tocify {
      width: 200px; /* Adjust this value to control TOC width */
    }

    /* Optional: Ensure content doesn't overlap with TOC on scroll */
    @media (min-width: 768px) { /* Apply for larger screens where TOC floats */
      body.toc-left .main-container {
        margin-left: 210px; /* Adjust this to be slightly more than TOC width */
      }
    }
    body {
  font-size: 12pt; /* Sets the base font size for the entire document */
}
h1,h2,h3,h4,h5,h6{
  font-size: 24pt;
}
code {
  font-size: 12pt; /* Sets font size for code blocks */
}
output {
  font-size: 24pt; /* Sets font size for code blocks */
}

</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center", fig.width = 12, fig.height = 6)
```

## Lab: Model Averaging using `mtcars` Data {.tabset}

In this lab, we will explore the concepts of model averaging. We'll use the `mtcars` dataset to predict miles per gallon (`mpg`), which is a regression task. We will build an "ensemble" of different models and show how a weighted average of their predictions can outperform any single model.

```{r load-packages}
rm(list = ls())
pacman::p_unload(pacman::p_loaded())
pacman::p_load(tidyverse, ISLR2, tree, patchwork, ggsci, caret, randomForest, gbm)

# Load the data
data(mtcars)
head(mtcars)
```

---

### Individual Models

We will create our "ensemble" by training four different types of models to predict `mpg`. We'll use 10-fold cross-validation (CV) to get reliable performance estimates (RMSE) for each one.

```{r}
set.seed(42)
# Define our training control (10-fold CV)
train_control <- trainControl(method = "cv", number = 10)

# Model 1: Simple Linear Regression (LM)
model_lm <- train(mpg ~ ., data = mtcars,
                  method = "lm", trControl = train_control)

# Model 2: LASSO Regression (a regularized model)
model_lasso <- train(mpg ~ ., data = mtcars,
                     method = "glmnet", trControl = train_control,
                     tuneGrid = expand.grid(alpha = 1, lambda = seq(0.01, 1, length = 100)))

# Model 3: K-Nearest Neighbors (KNN) (a non-linear model)
model_knn <- train(mpg ~ ., data = mtcars,
                   method = "knn", trControl = train_control,
                   tuneLength = 5)

# Model 4: Decision Tree (rpart)
model_tree <- train(mpg ~ ., data = mtcars,
                    method = "rpart", trControl = train_control,
                    tuneLength = 5)

# Collect the results
results <- resamples(list(LM = model_lm, LASSO = model_lasso, KNN = model_knn, TREE = model_tree))

# What is the "best" model from our CV
dotplot(results)
```

### Model averaging

```{r}
# Get predictions (Y-hat) from each model
predictions <- data.frame(
  truth = mtcars$mpg,
  lm = predict(model_lm, mtcars),
  lasso = predict(model_lasso, mtcars),
  knn = predict(model_knn, mtcars),
  tree = predict(model_tree, mtcars)
)
head(predictions)

# Average predictions - option 1: unweighted average
predictions$unweighted_average <- apply(predictions[,-1], 1, mean)

# Average predictions - option 2: weighted average. What weight to choose? Could use inverse RMSE (higher error predictions models get lower contribution)
weights <- 1/(results$values |> select(contains("RMSE")) |> colMeans()) ## average RMSE over all 10 folds.
weights <- weights/sum(weights) ## standardise weights so that they sum up to 1.

predictions$weighted_average <- as.matrix(predictions[,-c(1, 5)])%*%(weights)

RMSE_vals <- results$values |> select(contains("RMSE")) |> colMeans()
RMSE_vals <- c(RMSE_vals, `Unweighted~RMSE` = RMSE(predictions$unweighted_average, predictions$truth))
RMSE_vals <- c(RMSE_vals, `Weighted~RMSE` = RMSE(predictions$weighted_average, predictions$truth))

### print all RMSE values
RMSE_vals

as_tibble(predictions) |> 
  pivot_longer(cols = -truth) |> 
  ggplot(aes(x = truth, y = value, color = name)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  facet_grid(cols = vars(name)) + 
  theme_bw() + 
  theme(legend.position = "bottom") + 
  scale_color_aaas()
```
