---
title: "Class 7/Lab1: Clustering with Gaussian Mixture Models (GMMs)"
date: "Week 7 (10.17)"
output:
  html_document:
    toc: FALSE
    toc_float: true
    theme: united
---

```{=html}
<style type="text/css">
    .main-container {
      max-width: 1600px !important; /* Adjust this value as needed for the main content */
    }

    /* Optional: Adjust the width of the floating TOC */
    .tocify {
      width: 200px; /* Adjust this value to control TOC width */
    }

    /* Optional: Ensure content doesn't overlap with TOC on scroll */
    @media (min-width: 768px) { /* Apply for larger screens where TOC floats */
      body.toc-left .main-container {
        margin-left: 210px; /* Adjust this to be slightly more than TOC width */
      }
    }

</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

## Introduction {.tabset}

In this lab, we will use Gaussian Mixture Models (GMMs) to perform clustering on the `palmerpenguins` dataset. Our goal is to see if an unsupervised algorithm can discover the natural groupings (`species`) of penguins based on their physical measurements.



```{r load-packages}
# Clear environment and unload packages
rm(list = ls())
# The following line might show a note if no packages were loaded, which is normal.
pacman::p_unload(pacman::p_loaded())

# Load necessary packages
pacman::p_load(mclust, tidyverse, palmerpenguins, patchwork, cluster, here)
```

---

### Examine the Data

We load the `palmerpenguins` dataset. We will select only the numeric columns for our clustering analysis and remove any rows with missing data to ensure the algorithm runs smoothly.


```{r examine-data}
# Load the dataset
#penguins <- read_csv(file.path(here(), "class7", "data", "penguins.csv"))
penguins <- read_csv(file.path(dirname(here()), "data", "penguins.csv"))
# Inspect the data
head(penguins)
summary(penguins)

# Preprocess the data: select relevant columns and remove NAs
penguins_clean <- penguins %>%
  select(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g) %>%
  na.omit()

cat("Dimensions of the cleaned data:", dim(penguins_clean))
```


---

### Fitting the Gaussian Mixture Model

Now we'll fit the GMM using the `mclust` function. A key advantage of the `mclust` package is that it automatically evaluates different models and numbers of clusters (from 1 to 9 by default), then selects the best one using the Bayesian Information Criterion (`BIC`).

```{r visualize-data, fig.width=10, fig.height=5}
# Fit the GMM
gmm_model <- Mclust(penguins_clean)

# Print the summary of the best model
summary(gmm_model)
```

**Plot Interpretation:** The summary tells us that the best model found has 4 components (clusters). This is one more than the number of penguin species in the dataset!It also tells us the type of covariance structure it found (e.g., "VEE" which means `ellipsoidal`, `variable volume`, `equal shape`).

---

### Visualizing and Interpreting the Results

#### BIC Plot

First, let's look at the BIC values. The model with the highest BIC is considered the best. The plot clearly shows that the model with 4 clusters is the top-performing one.

```{r}
# Plot the BIC values for different models and numbers of clusters
plot(gmm_model, what = "BIC")
```


#### Classification Plot

Next, we can visualize the cluster assignments. The plot shows the data points colored by their assigned cluster, with ellipses representing the shape and orientation of each cluster's Gaussian distribution. We can see four groups. 

```{r}
# Plot the clustering results
plot(gmm_model, what = "classification")
```

####  Uncertainty Plot

We can also plot the model's uncertainty. Points with higher uncertainty (larger circles) are ones the model is less sure about, often lying on the boundary between two clusters.

```{r}
# Plot the uncertainty of classification
plot(gmm_model, what = "uncertainty")
```

#### Analyzing Cluster Characteristics

To understand what these clusters represent, we can add the cluster assignments back to our data and then calculate the average measurements for each cluster. This helps us build a "profile" for each group.

```{r}
# Add cluster assignments to the data
penguins_clean$cluster <- as.factor(gmm_model$classification)

# Group by cluster and calculate the mean of each feature
cluster_summary <- penguins_clean %>%
  group_by(cluster) %>%
  summarize(
    Avg_Bill_Length = mean(bill_length_mm),
    Avg_Bill_Depth = mean(bill_depth_mm),
    Avg_Flipper_Length = mean(flipper_length_mm),
    Avg_Body_Mass = mean(body_mass_g),
    Count = n()
  )

# Print the summary table
knitr::kable(cluster_summary, caption = "Mean Feature Values by GMM Cluster")
```

Interpretation: From the table, we can derive insights about physical differences between the penguin species. For example, one cluster might have penguins with long flippers and high body mass, while another has penguins with deep bills.

#### Stretch Goal: Comparison with True Labels (not always possible)

```{r}
# We need to remove the NAs from the original 'species' column in the same way we did for the numeric data.
penguins_with_species <- penguins %>%
  filter(row_number() %in% as.numeric(rownames(penguins_clean)))
# Create a confusion matrix
table(gmm_cluster = gmm_model$classification, true_species = penguins_with_species$species)
```






### Alternative: Finding Optimal Clusters with the Silhouette Score

Instead of relying only on the `BIC`, we can use the `silhouette score` to evaluate cluster quality. We will loop through several possible numbers of clusters (`k`), fit a GMM for each, and calculate the average silhouette score. The `k` with the highest score is considered optimal.

```{r}
# Calculate distance matrix once for efficiency
dist_matrix <- dist(penguins_clean)

# We will test k from 2 to 10
k_values <- 2:10
avg_sil_scores <- numeric(length(k_values))

for (i in 1:length(k_values)) {
  k <- k_values[i]
  
  # Fit GMM for a specific k
  gmm_fit <- Mclust(penguins_clean, G = k, verbose = FALSE)
  
  # If the model fails for a given k, it might return NULL classifications
  if(!is.null(gmm_fit)) {
    # Calculate silhouette score
    sil <- silhouette(gmm_fit$classification, dist_matrix)
    
    # Store the average silhouette width
    avg_sil_scores[i] <- mean(sil[, "sil_width"])
  } else {
    avg_sil_scores[i] <- NA # Assign NA if model failed
  }
}

# Create a data frame of the results
sil_results <- data.frame(k = k_values, avg_sil = avg_sil_scores)

# Plot the results
ggplot(sil_results, aes(x = k, y = avg_sil)) +
  geom_line() +
  geom_point() +
  labs(title = "Average Silhouette Score by Number of Clusters (k)",
       x = "Number of Clusters (k)",
       y = "Average Silhouette Score") +
  theme_minimal()

# Find the optimal k
optimal_k <- sil_results$k[which.max(sil_results$avg_sil)]
cat("\nOptimal number of clusters based on silhouette score is:", optimal_k, "\n")
```

Average silhouette score peaks at `k = 4` (agreeing with BIC), suggesting it's the best choice for our data. 
